---
title: "Final Project"
author: "Ava Mizrahi & Jake Whiting"
date: "2025-12-01"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Set Up 
```{r}
# reproducibility
set.seed(123)

# Load packages
library(tidyverse)
library(ggplot2)
library(dplyr)
library(car)
library(broom)
library(ggpubr)
```



Import datasets
```{r}
student_mat <- read.csv("student-mat.csv")
student_por <- read.csv("student-por.csv")

students <- bind_rows(student_mat %>% mutate(subject = "math"),
                      student_por %>% mutate(subject = "portuguese"))
```



Cleaning
```{r}
# Change variable names
students_clean <- students %>%
  select(subject, sex, age, internet, activities, studytime, absences, failures, goout, Dalc, Walc, G3) %>%
  rename(final_grade = G3) %>%
  mutate(sex = factor(sex, levels = c("F", "M"), labels = c("Female", "Male")), Dalc = as.numeric(Dalc), Walc = as.numeric(Walc))

head(students_clean)
str(students_clean)
summary(students_clean)
```



Exploratory Data Analysis
```{r}
# Histogram of final grades
ggplot(students_clean, aes(x = final_grade)) +
  geom_histogram(binwidth = 1, color = "white") +
  labs(title = "Distribution of Final Grades", x = "Final Grade", y = "Count") +
  theme_minimal()

# Histogram of Workday Alcohol Consumption
ggplot(students_clean, aes(x = Dalc)) +
  geom_histogram(binwidth = 1, color = "white") +
  labs(title = "Distribution of Workday Alc Consumption", x = "Dalc", y = "Count") +
  theme_minimal()

# Histogram of Weekend Alcohol Consumption
ggplot(students_clean, aes(x = Walc)) +
  geom_histogram(binwidth = 1, color = "white") +
  labs(title = "Distribution of Weekend Alc Consumption", x = "Walc", y = "Count") +
  theme_minimal()

# Boxplot of final grades by workday alcohol consumption
ggplot(students_clean, aes(x = factor(Dalc), y = final_grade, fill = factor(Dalc))) +
  geom_boxplot() +
  labs(title = "Final Grades by Workday Alcohol Consumption",
       x = "Dalc (1–5)", y = "Final Grade") +
  theme_minimal()

# Boxplot of final grades by weekend alcohol consumption
ggplot(students_clean, aes(x = factor(Walc), y = final_grade, fill = factor(Walc))) +
  geom_boxplot() +
  labs(title = "Final Grades by Weekend Alcohol Consumption",
       x = "Walc (1–5)", y = "Final Grade") +
  theme_minimal()

# Boxplot of final grades by going out
ggplot(students_clean, aes(x = factor(goout), y = final_grade, fill = factor(goout))) +
  geom_boxplot() +
  labs(title = "Final Grades by Go out",
       x = "Go Out (1–5)", y = "Final Grade") +
  theme_minimal()

# Scatterplot final grades vs. study time
ggplot(students_clean, aes(x = studytime, y = final_grade)) +
  geom_jitter(alpha = 0.3) +
  geom_smooth(method = "lm", color = "blue") +
  labs(title = "Final Grade vs Study Time",
       x = "Study Time (1-4)", y = "Final Grade")

# Boxplot of final grades by internet access
ggplot(students_clean, aes(x = factor(internet), y = final_grade, fill = factor(internet))) +
  geom_boxplot() +
  labs(title = "Final Grades by Internet",
       x = "Internet Access at Home", y = "Final Grade") +
  theme_minimal()

# Boxplot of final grades by activity
ggplot(students_clean, aes(x = factor(activities), y = final_grade, fill = factor(activities))) +
  geom_boxplot() +
  labs(title = "Final Grades by Activity",
       x = "Participation in Extracirriculars", y = "Final Grade") +
  theme_minimal()

# Scatterplot of final grades vs absences
ggplot(students_clean, aes(x = absences, y = final_grade)) +
  geom_jitter(alpha = 0.3) +
  geom_smooth(method = "lm", color = "blue") +
  labs(title = "Final Grade vs Absences",
       x = "Number of Absences", y = "Final Grade")

# Scatterplot of final grades vs number of past failures
ggplot(students_clean, aes(x = failures, y = final_grade)) +
  geom_jitter(alpha = 0.3) +
  geom_smooth(method = "lm", color = "blue") +
  labs(title = "Final Grade vs Failures",
       x = "Number of Failures", y = "Final Grade")
```


Two Factor ANOVA
```{r}
anova_md <- lm(final_grade ~ sex * Dalc, data = students_clean)
summary(anova_md)
Anova(anova_md, type = 2)
```

Logistic Regression Model
```{r}
# Make the final grade into a binary pass/fail variable with a 12 threshold
students_clean <- students_clean %>%
  mutate(pass = ifelse(final_grade >= 12, 1, 0),pass = factor(pass, levels = c(0,1), labels = c("Fail","Pass")))

table(students_clean$pass)

logit_mod <- glm(pass ~ sex + internet + failures + absences + activities + goout + studytime + Dalc + Walc , data = students_clean, family = binomial)

summary(logit_mod)
```



Odds Ratios
```{r}
# Odds ratios
exp(coef(logit_mod))
exp(confint(logit_mod))

logit_mod$null.deviance
logit_mod$deviance
1 - (logit_mod$deviance / logit_mod$null.deviance)
```



Prediction & Accuracy of Logistic Regression Model
```{r}
students_clean$pred_prob <- predict(logit_mod, type = "response")
students_clean$pred_class <- ifelse(students_clean$pred_prob > 0.6, "Pass", "Fail")

table(Predicted = students_clean$pred_class, Actual = students_clean$pass)

mean(students_clean$pred_class == students_clean$pass)
```


Fitted vs Observed
```{r}
ggplot(students_clean, aes(x = pass, y = pred_prob, fill = pass)) +
  geom_boxplot(alpha = 0.8) +
  labs(title = "Fitted Probabilities by Actual Outcome", x = "Actual Outcome", y = "Predicted Probability") +
  theme_minimal() +
  theme(legend.position = "none")
```


Reduced Logistic Model for Comparison
```{r}
# Top 4 significant variables
logit_reduced <- glm(pass ~ failures + absences + studytime + internet, data = students_clean, family = binomial)

summary(logit_reduced)

```

```{r}
# Compare AIC/BIC
AIC(logit_mod, logit_reduced)
BIC(logit_mod, logit_reduced)

anova(logit_reduced, logit_mod, test = "Chisq")
```


Parameter Estimation
```{r}
coef_table <- tidy(logit_mod, conf.int = TRUE, exponentiate = TRUE)
coef_table
```


Cook's Distance
```{r}
cooks <- cooks.distance(logit_mod)

plot(cooks, type = "h", main = "cook's Distance for logistic regression", ylab = "cook's distance")
```


Logistic Model Removing Influential Points
```{r}
influential_points <- which(cooks > (4 / nrow(students_clean)))

students_no_infl <- students_clean[-influential_points, ]
logit_noinfl <- glm(pass ~ sex + failures + absences + internet + activities + goout + studytime + Dalc + Walc, data = students_no_infl, family = binomial)

summary(logit_noinfl)
```

```{r}
# Compare models with full model and no influential points model 
AIC(logit_mod, logit_noinfl)
BIC(logit_mod, logit_noinfl)

logit_mod$deviance
logit_noinfl$deviance


```


Alcohol vs Subjects
```{r}
lm_subject_interaction <- lm(final_grade ~ subject * Dalc,
                             data = students_clean)

summary(lm_subject_interaction)
```


Absences vs Grade

```{r}
lm_absences <- lm(final_grade ~ absences + Dalc + sex + studytime, data = students_clean)
summary(lm_absences)
```




Male Logistic Model
```{r}
logit_male <- glm(pass ~ failures + internet + studytime + activities+ goout + absences + Dalc + Walc, data = filter(students_clean, sex == "Male"), family = binomial)
summary(logit_male)
```


Female Logistic Model
```{r}
logit_female <- glm(pass ~ failures + internet + activities + studytime + goout + absences + Dalc + Walc, data = filter(students_clean, sex == "Female"), family = binomial)
summary(logit_female)
```





